Stochastic Variance Reduced Policy Gradient

### 2018-01-30 svrpg.tex
added minibatch to algorithm description

\section{Algorithm}
\todomat{Add minibatch}
In on-line policy gradient problems, stochastic gradient ascent (SGA) is not really a choice, but is dictated by the necessity of interacting with an unknown environment. We would like to apply SVRG to the policy gradient framework in order to limit the variance introduced by sampling trajectories, which would ultimately lead to faster convergence. In this scenario, the function $f(\vtheta)$ to optimize is $J(\vtheta)$ and the (implicit) dataset $\mathcal{D}$ is the set of all possible trajectories $\mathcal{T}$. Compared to the typical finite sum optimization scenario, we have the following additional challenges:
\begin{itemize}
	\item The objective $J(\vtheta)$ is typically non-convex;
	\item The dataset $\mathcal{T}$ can be infinite;
	\item The value of the parameter $\vtheta$ influences the sampling of trajectories;
\end{itemize}
To deal with non-convexity, we require $J(\vtheta)$ to be L-smooth, which is a reasonable assumption for common policy classes such as truncated Gaussian and Softmax [appendix?]. Because of the infinite dataset, we can only rely on an estimate of the full gradient. Finally, we employ importance weighting \cite{rubinstein1981simulation} \cite{precup2000eligibility} to guarantee proper sampling of trajectories. We focus on the REINFORCE gradient estimator for simplicity, but the extension to G(PO)MDP is straightforward. The proposed SVRG-like gradient estimate, which we call Stochastic Variance Reduced Policy Gradient (SVRPG), is the following:
\begin{align*}
	\blacktriangledown J(\vtheta) &= \frac{1}{N}\sum_{j=0}^{N-1}\nabla\log\pi(\tau_j \mid \tilde{\vtheta})\Reward(\tau_j) \\
		&+ \nabla\log\pi(\tau_i \mid \vtheta)\Reward(\tau_i) 
		- \omega(\tau_i)\nabla\log\pi(\tau_i \mid \tilde{\vtheta})\Reward(\tau_i),
\end{align*}
where trajectories $\tau_j$ for $j=1,\dots,N$ are sampled from a snapshot policy $\pi_{\tilde{\vtheta}}$, $\tau_i$ is sampled from the current policy $\pol$, and $\omega(\tau) = \frac{\pi(\tau\mid\tilde{\vtheta})}{\pi(\tau\mid\vtheta)}$ is an importance weight from $\pol$ to the snapshot policy $\pi_{\tilde{\vtheta}}$. We immediately have the following (proof in Appendix \ref{app:proofs}):

\begin{restatable}{lemma}{unbias}\label{lemma:unbias}
The SVRPG gradient estimator is unbiased:
\[
\mathop{\mathbb{E}}
\left[\blacktriangledown J(\vtheta)\right] = \gradJ{\vtheta}.
\]
\end{restatable}
\todomat{Add lemma on zero variance at convergence}
The resulting SVRPG policy optimization method is detailed in Algorithm \ref{alg:svrpg}.
To obtain a method that can be used in practice, a number of additional details must be specified, namely the choice of epoch size $m$, the number $N$ of trajectories used to estimate the full gradient and the step size $\alpha$. Adaptive selection of these meta-parameters is also possible. However, the next section will focus on providing converge guarantees for the plain SVRPG algorithm.

\begin{algorithm}[tb]
	\caption{SVRPG}
	\label{alg:svrpg}
	\begin{algorithmic}
		\STATE {\bfseries Input:} number of epochs $S$, epoch size $m$, step size $\alpha$, initial parameter $\vtheta_{m-1}^0$, batch size $N$
		\FOR{$s=0$ {\bfseries to} $S-1$}
		\STATE Sample N trajectories $\tau_j$ from $\pi(\cdot\mid\tilde{\vtheta}^{s})$
		\STATE $\gradApp{\tilde{\vtheta}^{s}}{N} = \frac{1}{N}\sum_{j=0}^{N-1}\score{}{\tau_j\mid\tilde{\vtheta}^{s}}\Reward(\tau_j)$
		\STATE $\vtheta_0^{s+1} = \tilde{\vtheta}^s = \vtheta_{m-1}^s$
		\FOR{$t=0$ {\bfseries to} $m-1$}
		\STATE Sample trajectory $\tau_i$ from 				$\pi(\cdot\mid\vtheta_t^{s+1})$
		\STATE 
		\begin{align*}
		\blacktriangledown J(\vtheta_t^{s+1}) = 
		&\gradApp{\tilde{\vtheta}^s}{N}
		+ \score{}{\tau_i\mid\vtheta_t^{s+1}}\Reward(\tau_i) \\
		& - \omega(\tau_i)\score{}{\tau_i \mid \tilde{\vtheta}^{s}}\Reward(\tau_i)
		\end{align*}
		\STATE $\vtheta_{t+1}^{s+1} = \vtheta_t^{s+1} + \alpha\blacktriangledown J(\vtheta_t^{s+1})$
		\ENDFOR
		\ENDFOR
		\STATE {\bfseries return} $\vtheta_t^s$ with $(s,t)$ picked uniformly at random from $[0,S)\times[0,m)$
	\end{algorithmic}
\end{algorithm}

>>>>

\section{Algorithm}
\todomat{Add minibatch}
In on-line policy gradient problems, stochastic gradient ascent (SGA) is not really a choice, but is dictated by the necessity of interacting with an unknown environment. We would like to apply SVRG to the policy gradient framework in order to limit the variance introduced by sampling trajectories, which would ultimately lead to faster convergence. In this scenario, the function $f(\vtheta)$ to optimize is $J(\vtheta)$ and the (implicit) dataset $\mathcal{D}$ is the set of all possible trajectories $\mathcal{T}$. Compared to the typical finite sum optimization scenario, we have the following additional challenges:
\begin{itemize}
	\item The objective $J(\vtheta)$ is typically non-convex;
	\item The dataset $\mathcal{T}$ can be infinite;
	\item The value of the parameter $\vtheta$ influences the sampling of trajectories;
\end{itemize}
To deal with non-convexity, we require $J(\vtheta)$ to be L-smooth, which is a reasonable assumption for common policy classes such as truncated Gaussian and Softmax [appendix?]. Because of the infinite dataset, we can only rely on an estimate of the full gradient. Finally, we employ importance weighting \cite{rubinstein1981simulation} \cite{precup2000eligibility} to guarantee proper sampling of trajectories. 
We focus on the REINFORCE gradient estimator for simplicity, but the extension to G(PO)MDP is straightforward. The proposed SVRG-like gradient estimate, which we call Stochastic Variance Reduced Policy Gradient (SVRPG), is the following:
\begin{align*}
	\blacktriangledown J(\vtheta) &= \frac{1}{N}\sum_{j=0}^{N-1}\nabla\log\pi(\tau_j \mid \tilde{\vtheta})\Reward(\tau_j) \\
		&+ \nabla\log\pi(\tau_i \mid \vtheta)\Reward(\tau_i) 
		- \omega(\tau_i)\nabla\log\pi(\tau_i \mid \tilde{\vtheta})\Reward(\tau_i),
\end{align*}
where trajectories $\tau_j$ for $j=1,\dots,N$ are sampled from a snapshot policy $\pi_{\tilde{\vtheta}}$, $\tau_i$ is sampled from the current policy $\pol$, and $\omega(\tau) = \frac{\pi(\tau\mid\tilde{\vtheta})}{\pi(\tau\mid\vtheta)}$ is an importance weight from $\pol$ to the snapshot policy $\pi_{\tilde{\vtheta}}$. 
Our update is still fundamentally on-policy, since the weighting concerns only the correction term. However, this partial "off-policyness" is enough to introduce variance. To mitigate it, we use mini-batches of trajectories of size B to compute the second and the third term, as follows:
\begin{align*}
\blacktriangledown J(\vtheta) &= \frac{1}{N}\sum_{j=0}^{N-1}\nabla\log\pi(\tau_j \mid \tilde{\vtheta})\Reward(\tau_j) \\
&+ \frac{1}{B}\left[\sum_{i=0}^{B-1}
\nabla\log\pi(\tau_i \mid \vtheta)\Reward(\tau_i) 
- \omega(\tau_i)\nabla\log\pi(\tau_i \mid \tilde{\vtheta})\Reward(\tau_i)\right].
\end{align*}
The use of mini-batches is a common practice in SVRG since it can yield a performance improvement even in the supervised case \cite{konevcny2016mini} \cite{harikandeh2015stopwasting}.
We immediately have the following (proof in Appendix \ref{app:proofs}):
\begin{restatable}{lemma}{unbias}\label{lemma:unbias}
The SVRPG gradient estimator is unbiased:
\[
\mathop{\mathbb{E}}
\left[\blacktriangledown J(\vtheta)\right] = \gradJ{\vtheta}.
\]
\end{restatable}
\todomat{Add lemma on zero variance at convergence}
The resulting SVRPG policy optimization method is detailed in Algorithm \ref{alg:svrpg}.
To obtain a method that can be used in practice, a number of additional details must be specified, namely the choice of epoch size $m$, the number $N$ of trajectories used to estimate the full gradient, the mini-batch size $B$, and the step size $\alpha$. Adaptive selection of these meta-parameters is also possible. However, the next section will focus on providing general converge guarantees for Algorithm \ref{alg:svrpg}.

\begin{algorithm}[tb]
	\caption{SVRPG}
	\label{alg:svrpg}
	\begin{algorithmic}
		\STATE {\bfseries Input:} number of epochs $S$, epoch size $m$, step size $\alpha$, initial parameter $\vtheta_{m-1}^0$, batch size $N$, mini-batch size B
		\FOR{$s=0$ {\bfseries to} $S-1$}
		\STATE Sample N trajectories $\tau_j$ from $\pi(\cdot\mid\tilde{\vtheta}^{s})$
		\STATE $\gradApp{\tilde{\vtheta}^{s}}{N} = \frac{1}{N}\sum_{j=0}^{N-1}\score{}{\tau_j\mid\tilde{\vtheta}^{s}}\Reward(\tau_j)$
		\STATE $\vtheta_0^{s+1} = \tilde{\vtheta}^s = \vtheta_{m-1}^s$
		\FOR{$t=0$ {\bfseries to} $m-1$}
		\STATE Sample $B$ trajectories $\tau_i$ from 				$\pi(\cdot\mid\vtheta_t^{s+1})$
		\STATE 
		\begin{align*}
		\blacktriangledown J(\vtheta_t^{s+1}) = 
		&\gradApp{\tilde{\vtheta}^s}{N} \\
		&+\frac{1}{B}\sum_{i=0}^{B-1}\left[ 
		\score{}{\tau_i\mid\vtheta_t^{s+1}}\Reward(\tau_i)\right. \\
		&\left. - \omega(\tau_i)\score{}{\tau_i \mid \tilde{\vtheta}^{s}}\Reward(\tau_i)\right]
		\end{align*}
		\STATE $\vtheta_{t+1}^{s+1} = \vtheta_t^{s+1} + \alpha\blacktriangledown J(\vtheta_t^{s+1})$
		\ENDFOR
		\ENDFOR
		\STATE {\bfseries return} $\vtheta_t^s$ with $(s,t)$ picked uniformly at random from $[0,S)\times[0,m)$
	\end{algorithmic}
\end{algorithm}
